---
title: "Data Wrangling (1)"
author: "Haohan Chen"
date: "Last update: `r format(Sys.Date(), '%B %d, %Y')`"
output: 
  html_document: default
  pdf_document: default
  md_document: default
knit: (function(inputFile, encoding){rmarkdown::render(inputFile, encoding = encoding, output_format = "all", knit_root_dir = getwd())})
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Objectives of this Lecture

This lecture introduces data wrangling with R. Using V-Dem data as an example, we will learn how to use the wrangle data with a set of [`tidyverse`](https://www.tidyverse.org/) functionality. Specifically, we will focus on functions...

1.  to import and export data: `read_csv` , `write_csv` (with a brief introduction to other data import/ export functions from [`readr`](https://readr.tidyverse.org/)).

2.  to take a subset of *columns* in the existing data: `select`

3.  to rename columns: `rename`

4.  to take a subset of *rows* by some simple conditions: `slice_`

5.  to take a subset of *rows* by some more complicated conditions: `filter`

6.  to sort the rows based on the value of one or multiple columns: `arrange`

7.  to perform (4) (5) (6) group by group: `group_by`, `ungroup`

8.  to create new columns in the data: `group_by`, `mutate`, `ungroup`

9.  to summarize the data: `group_by`, `summarise`, `ungroup`

## Outline of In-Class Demo

To demonstrate the above functionality, we will use real-world political data from [V-Dem](https://v-dem.net/). Specifically, we will use the above function to explore the state of global economic development from 1984 to 2022. Our effort will take the following step (with one-on-one mappings with the above tools).

1.  Read a part of pre-processed V-Dem data into R: 1984-2022 "external" data in the V-Dem dataset.
2.  Consulting the dataset's [codebook](https://github.com/haohanchen/HKU_POLI3148_23Fall/blob/main/_DataPublic_/vdem/documentation/codebook_v13.pdf) and take a **subset** of indicators of *economic development* (along with country-year identifiers).
    -   *See a list of country-yer identifiers on p. 5 of the codebook (under "1.7 Identifier Variables in the V-Dem Datasets").*

    -   *See a list of development indicators on p. 23 of the codebook (under "9. Background Factors").*
3.  Rename the column to name their names informative to readers.
4.  Find the country-year with the *highest* and *lowest* level of economic development. In addition, create a dataset containing a random sample of country-year in the dataset.
5.  Create a dataset focusing on the economic development of Asian countries and regions; Create a dataset that contains only countries/ regions whose development level pass certain threshold.
6.  Create a dataset whose rows are sorted by the development level of country-year.
7.  Create a dataset that contains the year of the higest development level for each country/ region respectively.
8.  Add the following economic indicators to the data:
    1.  Country-year development level with reference to that of 1984.

    2.  Year-on-year economic growth.
9.  Perform a data availability/ integrity check. Then aggregate the data into a new country-level dataset which contains the following indicators:
    1.  Average development level from 1984 to 2022.

    2.  Magnitude of growth from 1984 to 2022.

## In-Class Exercise

The quality of education has a decisive effect on a country's future development. Applying the data wrangling tools we introduce in this lecture, perform the following task:

1.  **Coodbook lookup**. Look up the codebook, answer the following questions:
    1.  What indicators regarding the quality of education are available in the V-Dem datasets?

    2.  What are the data's coverage (i.e., for which countries and years do we have data?)

    3.  What are their sources? Provide the link to least 1 source.
2.  **Subset by columns**
    1.  Create a dataset containing only the country-year identifiers and indicators of education quality.

    2.  Rename the columns of education quality to make them informative.
3.  **Subset by rows**
    1.  List 5 countries-years that have the highest education level among its population.

    2.  List 5 countries-years that suffer from the most severe inequality in education.
4.  **Summarize the data**
    1.  Check data availability: For which countries and years are the indicators of education quality available?

    2.  Create two types of country-level indicators of education quality

        1.  Average level of education quality from 1984 to 2022

        2.  Change of education quality from 1984 to 2022

    3.  Examine the data and *briefly* discuss: Which countries perform the best and the worst in terms of education quality in the past four decades?

**Submission requirement:** You will submit your outputs through Moodle. In your submission:

1.  Attach a PDF document rendered by Rmarkdown
2.  In the text field of your submission, include the link to the corresponding Rmarkdown file in your *DaSPPA portfolio* GitHub repo.

**Due:** October 4, 2023

*Note:* *Please* *only use the functions we cover in this lecture for this exercise. There is [absolutely no need]{.underline} to perform any data visualization for this exercise... We will get there in later lectures.*

## Further reading

-   R for Data Science (2e) Chapters 4, 5, 8: <https://r4ds.hadley.nz/>

-   `readr` documentation (note: read the "cheatsheet"): <https://readr.tidyverse.org/>

-   `dplyr` documentation (note: read the "cheatsheet"): <https://dplyr.tidyverse.org/>

-   V-Dem documentation: <https://v-dem.net/>

## Demo

### 0. Load the `tidyverse` Packages

This section loads the packages we need in this lecture.

```{r}
library(tidyverse)
```

### 1. Import and Export the V-Dem Data

This section loads the VDEM dataset and describe its basic information

```{r}
d <- read_csv("_DataPublic_/vdem/1984_2022/vdem_1984_2022_external.csv")
getwd()
```

### 2. Select economic development indicators
```{r}
view(d)
```

First, we look at the identifiers of the data.
```{r}
names(d)
```

```{r}
d |> select(country_name, country_id, year) |> 
  distinct()
```

```{r}
# Which countries are in this dataset?
d |> select(country_name) |> 
  distinct()
```

Then, select both the country identifiers, GDP, and GDP per capita.

```{r}
d_gdp <- d |> select(country_name, country_id, year, e_gdp, e_gdppc)
d_gdp
```

###3. Rename columns to make names informative
```{r}
d_gdp <- d_gdp |>
  rename("GDP" = "e_gdp", "GDP_per_capita" = "e_gdppc", "Country" = "country_name", "ID" = "country_id", "Year" = "year")
view(d_gdp)
```

### 4. Slice rows
```{r}
#Want countries-years with highest GDP
d_gdp |> 
  slice_max(order_by = GDP, n = 10)
d_gdp |>
  slice_max(order_by = GDP, n = 5)
```

```{r}
d_gdp |>
  slice_min(order_by = GDP, n = 5)
```

```{r}
set.seed(52)
d_gdp |> slice_sample(n = 10)
```

```{r}
set.seed(52)
d_gdp |> slice_sample(prop = 0.1)
```

###5. Subset data by row
```{r}
#Want 2000-2005 data
d_gdp |> 
  filter(Year >= 2000 & Year <= 2005)
```

```{r}
d_gdp_china <- d_gdp |>
  filter(Country == "China")
view(d_gdp_china)
```

###6. Arrange
```{r}
#Want: sort the row by GDP per capita
d_gdp |> arrange(GDP_per_capita)
```

```{r}
d_gdp |> arrange(-GDP_per_capita)
d_gdp |> arrange((desc(GDP_per_capita)))
```

###7. Perform 4, 5, 6 group by group

```{r}
#Want: For each country, we want the year with the highest GDP
d_gdp |> 
  group_by(Country) |>
  slice_max(GDP, n = 1)
```

```{r}
#My attempt: 
#Want: For each year, we want the first 3 countries with the highest GDP
top_3_gdp <-
  d_gdp |>
  group_by(Year) |>
  slice_max(GDP, n = 3)
view(top_3_gdp) #SolvedQ: what about the extra lines? A: Those are NAs.
```

```{r}
#How many entries are there for each country?
d_gdp |>
  group_by(Country) |>
  count()
```

```{r}
#Want: The year when the countries have the worst GDP
d_gdp |>
  group_by(Country) |>
  slice_min(order_by = GDP, n = 1) #SolvedQ: Why is order_by needed? #A: No difference.
d_gdp |>
  group_by(Country) |>
  slice_min(GDP, n = 1)
?order_by
```

###8. Create new columns in the data: mutate

```{r}
d_gdp |>
  mutate (New = 1)
```

```{r}
d_gdp |>
  mutate (New = GDP)
```

```{r}
d_gdp |>
  mutate (New = log(GDP))
```

```{r}
d_gdp |>
  mutate (Pop = GDP / GDP_per_capita)
```

```{r}
d_gdp |>
  mutate (GDP_over_ave = GDP / mean(GDP, na.rm = TRUE))
#SolvedQ. Can we create only 1 new role? A. Nope.
?mutate

d_gdp |>
  group_by(Country) |>
  mutate (GDP_over_own_ave = GDP / mean(GDP, na.rm = TRUE))

#Want: Country-year development level with reference to 1984.
d_gdp |> 
  group_by(Country) |>
  arrange(Year) |>
  mutate (GDP_dev_level = GDP / first(GDP)) |>
  ungroup() |>
  arrange(Country, Year) #|>
  #print(n=20)
#The second arrange here is OK without Year.

#First function
?first

?lag

#Want: Year-on-year economic growth.
d_gdp |>
  group_by(Country) |>
  arrange(Year) |>
  mutate (GDP_year_dev = (GDP - lag(GDP, n = 1))/lag(GDP, n = 1)*100)|>
  ungroup() |>
  arrange(Year, Country)
#Q. Please make it prettier. Please use choose.
#A. For date of the year 1984, 
```

###9. Summarize the data: summarize

```{r}
#Want: Ave of world GDP.
d_gdp |>
  summarize(gdp_average = mean(GDP, na.rm = TRUE),
            gdp_per_capita_average = mean(GDP_per_capita, na.rm = TRUE))

#Want: Ave dev level.
d_gdp |>
  group_by(Country) |>
  summarize(GDP_average = mean(GDP, na.rm = TRUE),
            GDP_per_capita_average = mean(GDP_per_capita, na.rm = TRUE))

#Want: Overall GDP growth of 2019 over 1984.
d_gdp |>
  filter(Year >= 1984, Year <= 2019) |>
  group_by(Country) |>
  arrange(Year) |>
  summarize(GDP_growth_2019_1984 = (last(GDP) - first(GDP))/ first(GDP)) |>
  ungroup() |>
  arrange(Country, GDP_growth_2019_1984)

d_gdp |>
  filter(Year >= 1984, Year <= 2019) |>
  group_by(Country) |>
  arrange(Year) |>
  summarize(GDP_growth_percapita_2019_1984 = (last(GDP_per_capita) - first(GDP_per_capita))/ first(GDP_per_capita)) |>
  ungroup() |>
  arrange(Country, GDP_growth_percapita_2019_1984)

#Want: Data availability check.
d_gdp |>
  group_by(Country) |>
  mutate(GDP_missing = count(GDP))
?count
#Q. Why doesn't this work?

d_gdp |> 
  mutate(GDP_missing = as.numeric(is.na(GDP), .after = GDP)) |> #Put it after the GDP column.
  group_by(Country) |>
  summarize(N_GDP_missing = sum(GDP_missing))
```

###10. Pipe: insert left as right's first call.
?.data
```{r}
#Other options: see lecture notes.
#Shorten the code and maintain step-by-step flow.
#Avoid intermediate outputs. (Save environment, save time, save working space.)
#"|>" VS "%>%"
#Some other notes on "|>"
#Style: style.tidyverse.org
```

